Analyzing Inflation Dynamics in Canada During and After the COVID-19 Pandemic: A Deep Dive into Consumer Price Index Movements from 2019 to 2023

### 2024-04-04 outputs/paper.qmd
Update

---
title: "Analyzing Inflation Dynamics in Canada During and After the COVID-19 Pandemic: A Deep Dive into Consumer Price Index Movements from 2019 to 2023"
format: pdf
author: 
- Qin He
abstract: "The COVID-19 pandemic has left a significant imprint on global economic structures, influencing inflation rates worldwide. Canada, like many countries, experienced fluctuating inflation rates during and post-pandemic, marking a period of economic instability and adjustment. This paper delves into the inflation dynamics in Canada from 2019 to 2023, with a special focus on the consumer price index (CPI) as a reflection of the cost of goods and services impacting the everyday lives of Canadians. It explores the trajectory of the overall inflation rate, which peaked in 2023, highlighting the role of various sectors, including food, housing, and energy, in driving these trends. By analyzing data from Statistics Canada, this study provides insights into how the pandemic and subsequent recovery phases have affected inflation. It examines the interplay between supply chain disruptions, monetary policy responses, and changes in consumer behavior, culminating in a comprehensive overview of inflation's impact on the Canadian economy. This analysis not only sheds light on the inflationary trends but also contributes to understanding the broader economic implications of the COVID-19 pandemic on Canada."
date: today
date-format: long
editor: visual
toc: TRUE
toc-depth: 1
number-sections: TRUE
bibliography: references.bib
thanks: "See the GitHub repository supporting this analysis: https://github.com/qinheinfo/Canada_inflation"
---

@@

# Fetch the the clean inflation data
inflation_data <- arrow::read_parquet(here::here("----"), show_col_types = FALSE)

# Fetch clean CPI grocery item data
grocery_data <- arrow::read_parquet(here::here("----"), show_col_types = FALSE)
```

```

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

---
title: "Analyzing Inflation Dynamics in Canada During and After the COVID-19 Pandemic: A Deep Dive into Consumer Price Index Movements from 2019 to 2023"
format: html
author: 
- Qin He
abstract: "The COVID-19 pandemic has left a significant imprint on global economic structures, influencing inflation rates worldwide. Canada, like many countries, experienced fluctuating inflation rates during and post-pandemic, marking a period of economic instability and adjustment. This paper delves into the inflation dynamics in Canada from 2019 to 2023, with a special focus on the consumer price index (CPI) as a reflection of the cost of goods and services impacting the everyday lives of Canadians. It explores the trajectory of the overall inflation rate, which peaked in 2023, highlighting the role of various sectors, including food, housing, and energy, in driving these trends. By analyzing data from Statistics Canada, this study provides insights into how the pandemic and subsequent recovery phases have affected inflation. It examines the interplay between supply chain disruptions, monetary policy responses, and changes in consumer behavior, culminating in a comprehensive overview of inflation's impact on the Canadian economy. This analysis not only sheds light on the inflationary trends but also contributes to understanding the broader economic implications of the COVID-19 pandemic on Canada."
date: today
date-format: long
editor: visual
toc: TRUE
toc-depth: 1
number-sections: TRUE
bibliography: references.bib
thanks: "See the GitHub repository supporting this analysis: https://github.com/qinheinfo/Canada_inflation"
---

@@

# Fetch clean CPI product item data
product_data <- read.csv(here::here("data/cleaned_data.csv"), stringsAsFactors = FALSE)
```

@@

## Food

```{r}
#| messages: false
#| echo: false
#| warning: false

# Load necessary libraries
library(ggplot2)
library(readr)

# Read the dataset
data <- read.csv(here::here("data/cleaned_data.csv"))

# Filter for the "Food" and "All-items" product groups
filtered_data <- data[data$products_and_product_groups %in% c('Food', 'All-items'),]

# Plotting
ggplot(filtered_data, aes(x = ref_date, y = value, color = products_and_product_groups, group = products_and_product_groups)) +
  geom_line(size = 1) +  # Draw lines for both groups
  geom_point(size = 2) +  # Highlight data points for both groups
  labs(
    title = "Trend of Value of Food and All-items from 2019 to 2023",
    y = "Value",
    x = "Year",
    color = "Product Group"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = filtered_data$ref_date) # Ensure all years are shown

```
```{r}
#| messages: false
#| echo: false
#| warning: false
#| label: fig-2
#| fig-cap: "Comparison of Food Value and All-items Value Over Time"
#| fig-align: center
#| fig-width: 10
#| fig-height: 3
# Read and prepare the dataset for displaying in a table format
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)

# Read the dataset
data <- read.csv(here::here("data/cleaned_data.csv"))

# Reshape data to wide format using modern pipe syntax
wide_data <- data |> 
  filter(products_and_product_groups %in% c('Food', 'All-items')) |>
  spread(key = products_and_product_groups, value = value) |>
  rename(Year = ref_date)

# Assuming that the 'value' columns are already in the correct format and do not need additional formatting

# Prepare the data for the table display
wide_data_display <- wide_data |>
  select(Year, Food, `All-items`) |> 
  rename(
    `Food Value` = Food, 
    `All-items Value` = `All-items`
  )

# Render the table using knitr::kable and kableExtra::kable_styling
wide_data_display |>
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "" ) |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE)
```

@@

@@
